---
title: "Portrait of Welfare Recipients"
description: |
  Some descriptive statistics about recipients of welfare.
author:
  - name: Ornella Darova
  - name: Ozgur Seker
date: 03-10-2021
output:
  distill::distill_article:
    self_contained: false
categories: Ornella and Ozgur's project
---

# Data

We have a rich dataset from the CPS IPUMS that has observations collected from 2009 to 2020 and has 183 variables. We are going to complement this dataset with additional information at the county or state level, as it might be relevant for the analysis we plan to conduct.

# Linking the Census Data to County Level Data

Our goal in this post is to conduct an exploratory analysis of our data, and moreover to add other sources of information to the main CPS dataset.

We start by the second task.

```{r setup, include=TRUE}
library(tidyverse)
library(haven)
library(janitor)
library(expss)

readCountyData <- function(){
  countyData <- read_csv("~/Dropbox/Penn/ML/Predicting Take-Up Rate/CountyLevelData/CAINC1/CAINC1__ALL_AREAS_1969_2019.csv")
  countyData = countyData[1:9528,]
  return(countyData)
}


cd <- readCountyData()
cols <- names(cd)[9:length(names(cd))]
cd <- cd %>% mutate_at(cols, as.numeric)

prepCountyData <- function(data, year = 2019, newColNames = c("Income", "Pop", "PerCapInc"), colsNeeded = c("GeoFIPS", "GeoName", "Region") ){
  
  groups <- unique(data$LineCode)
  dataNew <- data %>% filter(LineCode == groups[1])
  dataNew[, paste0(year,"_",newColNames[1])] <- dataNew[, paste(year)]
  
  yourCols <- c(colsNeeded, paste0(year,"_",newColNames[1]))
  
  for (i in 2:length(groups)){
    grp <- groups[i]
    colnam <- newColNames[i]
    dataTemp <- data %>% filter(LineCode == grp) 
    newName <- paste0(year,"_",newColNames[i]) 
    yourCols <- c(yourCols, newName)
    dataNew[,newName] <- dataTemp[,paste(year)]
  }
  
  return(dataNew[,yourCols])
  
}

cd <- prepCountyData(cd)
cd$GeoFIPS  <- as.numeric(cd$GeoFIPS)

head(cd)
```

# Explorative Analysis of the Dataset

We now proceed with the explorative analysis of the main dataset. 

```{r, include=TRUE,echo=FALSE}

CreateMissingDataFrame <- function(){
  MissingValsDF <- data.frame( Var = "faminc", Val = 995 )
  MissingValsDF <- MissingValsDF %>% add_row(Var = "faminc", Val = 997)
  MissingValsDF <- MissingValsDF %>% add_row(Var = "metarea", Val = 9999)
  MissingValsDF <- MissingValsDF %>% add_row(Var = "statefip", Val = 99)
  MissingValsDF <- MissingValsDF %>% add_row(Var = "faminc", Val = 999)
  MissingValsDF <- MissingValsDF %>% add_row(Var = "race", Val = 999)
  MissingValsDF <- MissingValsDF %>% add_row(Var = "famsize", Val = 0)
  MissingValsDF <- MissingValsDF %>% add_row(Var = "mbpl", Val = 96000)
  MissingValsDF <- MissingValsDF %>% add_row(Var = "nativity", Val = 0)
  MissingValsDF <- MissingValsDF %>% add_row(Var = "hispan", Val = 901)
  MissingValsDF <- MissingValsDF %>% add_row(Var = "uhrsworkt", Val = 997)
  MissingValsDF <- MissingValsDF %>% add_row(Var = "wnlook", Val = 997)
  MissingValsDF <- MissingValsDF %>% add_row(Var = "migrate1", Val = 9)
  MissingValsDF <- MissingValsDF %>% add_row(Var = "paidhour", Val = 7)
  write_csv(MissingValsDF, file = "missingValues.csv")
}

CreateRefusedDataFrame <- function(){
  RefusadValsDF <- data.frame( Var = "faminc", Val = 996 )
  RefusadValsDF <- RefusadValsDF %>% add_row(Var = "empstat", Val = 0)
  RefusadValsDF <- RefusadValsDF %>% add_row(Var = "paidhour", Val = 6)
  write_csv(RefusadValsDF, file = "refusedValues.csv")
}

CreateNiuDataFrame <- function(){
  NiuValsDF <- data.frame( Var = "vetstat", Val = 0 )
  NiuValsDF <- NiuValsDF %>% add_row(Var = "empstat", Val = 0)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "metarea", Val = 9998)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "ownershp", Val = 0)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "pubhous", Val = 0)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "rentsub", Val = 0)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "heatsub", Val = 0)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "foodstmp", Val = 0)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "stmpmo", Val = 0)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "atelunch", Val = 99)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "empstat", Val = 0)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "lunchsub", Val = 0)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "frelunch", Val = 98)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "frelunch", Val = 99)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "unitsstr", Val = 0)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "phone", Val = 0)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "sex", Val = 9)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "marst", Val = 9)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "eldch", Val = 99)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "yngch", Val = 9)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "bpl", Val = 99999)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "yrimmig", Val = 0)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "citizen", Val = 9)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "mbpl", Val = 99999)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "labforce", Val = 0)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "uhrsworkt", Val = 999)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "durunem2", Val = 99)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "whyunemp", Val = 0)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "wnlook", Val = 999)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "wkstat", Val = 99)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "educ", Val = 1)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "educ", Val = 0)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "diffany", Val = 0)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "pension", Val = 0)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "firmsize", Val = 0)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "payifabs", Val = 0)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "srceduc", Val = 0)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "srcunemp", Val = 9)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "srcwelfr", Val = 0)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "srcwkcom", Val = 0)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "ssikid", Val = 0)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "sskid", Val = 0)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "whyss1", Val = 0)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "whyss2", Val = 0)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "whyssi1", Val = 0)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "whyssi2", Val = 0)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "gotvdisa", Val = 0)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "gotveduc", Val = 0)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "gotvothe", Val = 0)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "gotvpens", Val = 0)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "gotsurv", Val = 0)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "migsta1", Val = 0)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "migsta5", Val = 0)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "whymove", Val = 0)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "migrate1", Val = 0)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "disabwrk", Val = 0)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "quitsick", Val = 0)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "inclugh", Val = 0)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "paidgh", Val = 0)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "hichamp", Val = 9)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "phinsur", Val = 0)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "phiown", Val = 0)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "caidly", Val = 9)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "carely", Val = 9)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "schiply", Val = 0)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "hinelig1", Val = 99)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "hinelig2", Val = 99)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "hinelig3", Val = 99)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "hinelig4", Val = 99)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "hinelig5", Val = 99)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "kidpriv", Val = 0)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "kidcaid", Val = 0)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "gotwic", Val = 0)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "paidhour", Val = 0)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "union", Val = 0)
  NiuValsDF <- NiuValsDF %>% add_row(Var = "heatval", Val = 0)
  
  write_csv(NiuValsDF, file = "NIUValues.csv")
}

# CreateMissingDataFrame()
# CreateRefusedDataFrame()
# CreateNiuDataFrame()
```

```{r, include=TRUE}

cpsSample <- read_dta("~/Dropbox/Penn/ML/Predicting Take-Up Rate/sampleData.dta")

missingDF <- read_csv(file = "~/Dropbox/Penn/ML/Predicting Take-Up Rate/missingValues.csv")
niuDF <- read_csv(file = "~/Dropbox/Penn/ML/Predicting Take-Up Rate/NIUValues.csv")
refusedDF <- read_csv(file = "~/Dropbox/Penn/ML/Predicting Take-Up Rate/refusedValues.csv")


orderedVars <- c("heatval", "stampmo", "atelunch", "frelunch", "faminc", "unitsstr", 
                 "age", "famsize", "nchild", "nchlt5", "famunit", "eldch", "yngch", "durunem2",
                 "uhrsworkt", "educ", "firmsize", "health")

ncps <- cpsSample %>% left_join(cd%>%select(GeoFIPS, "2019_Income", "2019_Pop", "2019_PerCapInc"), by= c("county" = "GeoFIPS"))

cor(ncps$health, ncps$hhincome, use = "complete.obs")
cor(ncps$health, ncps$faminc, use = "complete.obs")
cor(ncps$health, ncps$hhincome, use = "complete.obs")

#ncps <- ncps %>% filter(county != 0)
#cor(ncps$famsize, ncps$`2019_PerCapInc`, use = "complete.obs")
#cor(ncps$hhincome, ncps$`2019_PerCapInc`, use = "complete.obs")
#cor(ncps$heatsub, ncps$`2019_PerCapInc`, use = "complete.obs")
#cor(ncps$health, ncps$`2019_PerCapInc`, use = "complete.obs")

# rm(cd)
# rm(cpsSample)
# rm(cols)
# rm(prepCountyData)
# rm(readCountyData)

```

We now analyze the take-up rate of different social programs among individuals living in household units below the Supplemental Poverty Measure.

First of all, we try to understand, among those who are deemed to be in the universe of possible participants to social welfare, how many actually participate.

```{r, include=TRUE,echo=FALSE}

library(pdp)
library(ggplot2)

YesNo <- c("Yes", "No")
cpsSamplePov <- subset(cpsSample, cpsSample$spmpov==1)

p1 <- ggplot(as.data.frame(subset(cpsSamplePov,cpsSamplePov$pubhous!=0)), 
       aes(x=as.factor(subset(cpsSamplePov$pubhous,cpsSamplePov$pubhous!=0)))) + 
       geom_bar(alpha = .2, aes(fill=..count..)) +
       scale_fill_gradient("count", low="green", high="red") + 
       scale_x_discrete(labels= YesNo) + 
       ylab("Living in Public Housing") + 
       theme(legend.position="none", axis.title.x=element_blank())

p2 <- ggplot(as.data.frame(subset(cpsSamplePov,cpsSamplePov$rentsub!=0)), 
       aes(x=as.factor(subset(cpsSamplePov$rentsub,cpsSamplePov$rentsub!=0)))) + 
       geom_bar(alpha = .2, aes(fill=..count..)) +
       scale_fill_gradient("count", low="green", high="red") + 
       scale_x_discrete(labels= YesNo) + 
       ylab("Government Rent Subsidy Recipient") + 
       theme(legend.position="none", axis.title.x=element_blank())

p3 <- ggplot(as.data.frame(subset(cpsSamplePov,cpsSamplePov$heatsub!=0)), 
       aes(x=as.factor(subset(cpsSamplePov$heatsub,cpsSamplePov$heatsub!=0)))) + 
       geom_bar(alpha = .2, aes(fill=..count..)) +
       scale_fill_gradient("count", low="green", high="red") + 
       scale_x_discrete(labels= YesNo) + 
       ylab("Government Heat Subsidy Recipient") + 
       theme(legend.position="none", axis.title.x=element_blank())

p4 <- ggplot(as.data.frame(subset(cpsSamplePov,cpsSamplePov$foodstmp!=0)), 
       aes(x=as.factor(subset(cpsSamplePov$foodstmp,cpsSamplePov$foodstmp!=0)))) + 
       geom_bar(alpha = .2, aes(fill=..count..)) +
       scale_fill_gradient("count", low="green", high="red") + 
       scale_x_discrete(labels= YesNo) + 
       ylab("Government Food Stamps Recipient") + 
       theme(legend.position="none", axis.title.x=element_blank())

p5 <- ggplot(as.data.frame(subset(cpsSamplePov,cpsSamplePov$lunchsub!=0)), 
       aes(x=as.factor(subset(cpsSamplePov$lunchsub,cpsSamplePov$lunchsub!=0)))) + 
       geom_bar(alpha = .2, aes(fill=..count..)) +
       scale_fill_gradient("count", low="green", high="red") + 
       scale_x_discrete(labels= YesNo) + 
       ylab("Government Lunch Subsidy Recipient") + 
       theme(legend.position="none", axis.title.x=element_blank())

grid.arrange(p1,p2,nrow=1)
grid.arrange(p3,p4,nrow=1)
p5
```

Subsequently, we are going to try to understand, keeping our discussion around individuals whose household is below the poverty line, what is the distribution of income according to the participation to the previously discussed social programs.

```{r, include=TRUE,echo=FALSE}
p1 <- ggplot(as.data.frame(cpsSamplePov), aes(x=as.factor(cpsSamplePov$faminc))) +
      geom_bar(aes(fill=as.factor(cpsSamplePov$pubhous)), 
      position = "identity", alpha=0.3) + theme(axis.text.x = element_blank()) + 
      ylab("Living in Public Housing") + xlab("Household Income") + 
      scale_fill_discrete(name = "Public Housing", labels = c("Not in Universe", "No", "Yes"))

p2 <- ggplot(as.data.frame(cpsSamplePov), aes(x=as.factor(cpsSamplePov$faminc))) +
      geom_bar(aes(fill=as.factor(cpsSamplePov$rentsub)), 
      position = "identity", alpha=0.3) + theme(axis.text.x = element_blank()) + 
      ylab("Government Rent Subsidy Recipient") + xlab("Household Income") + 
      scale_fill_discrete(name = "Rent Subsidy Recipient", labels = c("Not in Universe", "No", "Yes"))

p3 <- ggplot(as.data.frame(cpsSamplePov), aes(x=as.factor(cpsSamplePov$faminc))) +
      geom_bar(aes(fill=as.factor(cpsSamplePov$heatsub)), 
      position = "identity", alpha=0.3) + theme(axis.text.x = element_blank()) + 
      ylab("Government Heat Subsidy Recipient") + xlab("Household Income") + 
      scale_fill_discrete(name = "Heat Subsidy Recipient", labels = c("Not in Universe", "No", "Yes"))

p4 <- ggplot(as.data.frame(cpsSamplePov), aes(x=as.factor(cpsSamplePov$faminc))) +
      geom_bar(aes(fill=as.factor(cpsSamplePov$foodstmp)), 
      position = "identity", alpha=0.3) + theme(axis.text.x = element_blank()) + 
      ylab("Government Food Stamps Recipient") + xlab("Household Income") + 
      scale_fill_discrete(name = "Food Stamps Recipient", labels = c("Not in Universe", "No", "Yes"))

p5 <- ggplot(as.data.frame(cpsSamplePov), aes(x=as.factor(cpsSamplePov$faminc))) +
      geom_bar(aes(fill=as.factor(cpsSamplePov$lunchsub)), 
      position = "identity", alpha=0.3) + theme(axis.text.x = element_blank()) + 
      ylab("Government Lunch Subsidy Recipient") + xlab("Household Income") + 
      scale_fill_discrete(name = "Lunch Subsidy Recipient", labels = c("Not in Universe", "No", "Yes"))

p1
p2
p3
p4
p5

```

We are now going to explore the ethnic distribution of those who apply to social programs.

```{r, include=TRUE,echo=FALSE}

cpsSamplePovRace <- subset(cpsSamplePov,race==100|race==200|race==300|race==651)
Race <- c("White","Black","American Indian","Asian")

p1 <- ggplot(as.data.frame(cpsSamplePovRace), aes(x=as.factor(cpsSamplePovRace$race))) +
      geom_bar(aes(fill=as.factor(cpsSamplePovRace$pubhous)), 
      position = "identity", alpha=0.3) + scale_x_discrete(labels= Race) +
      ylab("Living in Public Housing") + xlab("Race") + 
      scale_fill_discrete(name = "Public Housing", labels = c("Not in Universe", "No", "Yes"))

p2 <- ggplot(as.data.frame(cpsSamplePovRace), aes(x=as.factor(cpsSamplePovRace$race))) +
      geom_bar(aes(fill=as.factor(cpsSamplePovRace$rentsub)), 
      position = "identity", alpha=0.3) + scale_x_discrete(labels= Race) +
      ylab("Government Rent Subsidy Recipient") + xlab("Race") + 
      scale_fill_discrete(name = "Rent Subsidy Recipient", labels = c("Not in Universe", "No", "Yes"))

p5 <- ggplot(as.data.frame(cpsSamplePovRace), aes(x=as.factor(cpsSamplePovRace$race))) +
      geom_bar(aes(fill=as.factor(cpsSamplePovRace$lunchsub)), 
      position = "identity", alpha=0.3) + scale_x_discrete(labels= Race) +
      ylab("Government Lunch Subsidy Recipient") + xlab("Race") + 
      scale_fill_discrete(name = "Lunch Subsidy Recipient", labels = c("Not in Universe", "No", "Yes"))

p1
p2
p3
p4
p5
```


