---
title: "Clustering and PCA Analysis"
description: |
  An analysis of Medicare take-up and a Singular value decomposition exercise
author:
  - name: Ornella Darova
  - name: Ozgur Seker
    url: {}
date: "`r Sys.Date()`"
output: distill::distill_article
categories: Ornella and Ozgur's project
---

# Clustering

## Medicare

We perform a clustering exercise at the state level, focusing on the take-up of the Medicare program.
```{r prep, echo=FALSE,warnings=FALSE, message=FALSE,include=FALSE}
library(dplyr)
library(ranger)   
library(rsample)
library(caret)
library(rpart)
library(haven)
library(rlang)
library(tibble)

ncps <- read_dta("/Users/ornelladarova/Dropbox/Penn/ML/Predicting Take-Up Rate/sampleDataFull.dta")

dataPrep <- function(ncps = ncps){

  
  ### Build eligibility variable for PUBLIC HOUSING
  ncps <- as.data.frame(ncps)
  ncps$hous_elig = ifelse(ncps$famincAssigned<ncps$l80|ncps$disabwrk==2|ncps$age>65,1,0) 
  ncps$hous_elig = ifelse(ncps$disabwrk==0,".",ncps$hous_elig)
  ncps$hous_elig = as.numeric(ncps$hous_elig)
  
  ### Build eligibility variable for MEDICARE
  
  ncps$med_elig = ifelse(ncps$disabwrk==2|ncps$age>65,1,0)
  ncps$med_elig = ifelse(ncps$disabwrk==0,".",ncps$med_elig)
  ncps$med_elig = as.numeric(ncps$med_elig)
  
  
  # Prepare variables
  ncps$pub <- ifelse(ncps$pubhous==1,0,1)
  ncps$pub <- ifelse(ncps$pubhous==0,".",ncps$pub)
  ncps$pub <- as.numeric(ncps$pub)
  
  ncps$medicare <- ifelse(ncps$carely==1,0,1)
  ncps$medicare <- ifelse(ncps$carely==0,".",ncps$medicare)
  ncps$medicare <- as.numeric(ncps$medicare)
  
  ncps$age <- as.numeric(ncps$age)
  ncps$male <- ifelse(ncps$sex==2,0,1)
  
  ncps$male <- ifelse(ncps$sex==9,".",ncps$male)
  ncps$male <- as.numeric(ncps$male)
  
  ncps$black <- ifelse(ncps$race==200,1,".")
  ncps$black <- ifelse(ncps$race==100,0,ncps$black)
  ncps$black <- as.numeric(ncps$black)
  ncps$asian <- ifelse(ncps$race==651,1,0)
  ncps$asian <- as.numeric(ncps$asian)
  
  ncps$married <- ifelse(ncps$marst==1|ncps$marst==2,1,0)
  ncps$married <- ifelse(ncps$marst==9,".",ncps$married)
  ncps$married <- as.numeric(ncps$married)
  
  ncps$foreign <- ifelse(ncps$nativity==4|ncps$nativity==5,1,".")
  ncps$foreign <- ifelse(ncps$nativity==1,0,ncps$foreign)
  ncps$foreign <- as.numeric(ncps$foreign)
  
  ncps$veteran <- ifelse(ncps$vetstat==1,0,".")
  ncps$veteran <- ifelse(ncps$vetstat==2,1,ncps$veteran)
  ncps$veteran <- as.numeric(ncps$veteran)
  
  ncps$nchildren <- as.numeric(ncps$nchild)
  
  ncps$unemployed <- ifelse(ncps$labforce==2,0,1)
  ncps$unemployed <- ifelse(ncps$labforce==0,".",ncps$unemployed)
  ncps$unemployed <- as.numeric(ncps$unemployed)
  
  ncps$phone <- ifelse(ncps$phone==0|ncps$phone==9,NA,ncps$phone)
  ncps$phone <- ifelse(ncps$phone==1,0,ncps$phone)
  ncps$phone <- ifelse(ncps$phone==2|ncps$phone==3,1,ncps$phone)
  
  return(ncps)
}

ncps <- dataPrep(ncps)


```

```{r clusters, echo=TRUE,warnings=FALSE,message=FALSE}
library(cluster)
library(factoextra)

# Consider eligible individuals for medicare
ncps_el = subset(ncps, ncps$med_elig==1)
ncps_sub = select(ncps_el,c("famincAssigned","pub","medicare","male","black",
               "married","foreign","veteran","nchildren","unemployed",
               "phone","is_largefmly","health","statefip"))
ncps_sub[c("famincAssigned","pub","medicare","male","black",
               "married","foreign","veteran","nchildren","unemployed",
               "phone","is_largefmly","health")] = ncps_sub[c("famincAssigned","pub","medicare","male","black",
               "married","foreign","veteran","nchildren","unemployed",
               "phone","is_largefmly","health")] %>% mutate_all(as.numeric)

```

```{r clusters_1, echo=TRUE,warnings=FALSE,message=FALSE}
# Consider first those who actually take up the program
take_up = subset(ncps_sub,medicare==1)
take_up = take_up[c("famincAssigned","pub","male","black",
               "married","foreign","veteran","nchildren","unemployed",
               "phone","is_largefmly","health","statefip")]

# Aggregate at the state level
ncps_statefip <- aggregate(take_up, by=list(take_up$statefip), FUN=mean, na.rm = TRUE)
ncps_statefip$famincAssigned <- scale(ncps_statefip$famincAssigned)

## Elbow method to find optimal number of clusters
set.seed(123)

# Function to compute total within-cluster sum of square 
wss <- function(k) {
  kmeans(na.omit(ncps_statefip), k, nstart = 10 )$tot.withinss
}

# Compute wss
k.values <- 1:10

# Extract wss for different numbers of clusters
library(purrr)
wss_values <- purrr::map_dbl(k.values, wss)

plot(k.values, wss_values,
       type="b", pch = 19, frame = FALSE, 
       xlab="Number of clusters K",
       ylab="Total within-clusters sum of squares")

## 4 clusters seems to be sufficient
# Clustering

rownames(ncps_statefip) = ncps_statefip$statefip

ncps_statefip = select(ncps_statefip,c("famincAssigned","pub","male","black",
               "married","foreign","veteran","nchildren","unemployed",
               "phone","is_largefmly","health"))
k2 <- kmeans(na.omit(ncps_statefip), centers = 4, nstart = 25)
str(k2)


fviz_cluster(k2, data = na.omit(ncps_statefip))

na.omit(ncps_statefip) %>%
  mutate(Cluster = k2$cluster) %>%
  group_by(Cluster) %>%
  summarise_all("mean")

na.omit(ncps_statefip) %>%
  mutate(Cluster = k2$cluster) %>%
  group_by(Cluster)

```

The formed clusters are the following:

1) Connecticut, District of Columbia, Hawaii, Marayland, New Hampshire, New Jersey, Virginia, Washington
2) Arizona, Delaware, Georgia, Illinoi, Indiana, Iowa, Maine, Michigan, Minnesota, Nebraska, Nevada, New Mexico, Oklahoma, Oregon, Pennsylvania, Tennessee
3) California, Colorado, Kansas, Massachussetts, Missouri, New York, Utah
4) Alabama, Arkansas, Florida, Idaho, Kentucky, Louisiana, Montana, North Carolina, North Dakota, Ohio, South Carolina, Texas, West Virginia, Wisconsin


```{r clusters_2, echo=FALSE,warnings=FALSE,message=FALSE}
# Consider now those who do not take up the program
take_up = subset(ncps_sub,medicare==0)
take_up = take_up[c("famincAssigned","pub","male","black",
               "married","foreign","veteran","nchildren","unemployed",
               "phone","is_largefmly","health","statefip")]

# Aggregate at the state level
ncps_statefip <- aggregate(take_up, by=list(take_up$statefip), FUN=mean, na.rm = TRUE)
ncps_statefip$famincAssigned <- scale(ncps_statefip$famincAssigned)

## Elbow method to find optimal number of clusters
set.seed(123)

# Function to compute total within-cluster sum of square 
wss <- function(k) {
  kmeans(na.omit(ncps_statefip), k, nstart = 10 )$tot.withinss
}

# Compute wss
k.values <- 1:10

# Extract wss for different numbers of clusters
library(purrr)
wss_values <- purrr::map_dbl(k.values, wss)

plot(k.values, wss_values,
       type="b", pch = 19, frame = FALSE, 
       xlab="Number of clusters K",
       ylab="Total within-clusters sum of squares")

## 4 clusters seems to be sufficient
# Clustering

rownames(ncps_statefip) = ncps_statefip$statefip

ncps_statefip = select(ncps_statefip,c("famincAssigned","pub","male","black",
               "married","foreign","veteran","nchildren","unemployed",
               "phone","is_largefmly","health"))
k2 <- kmeans(na.omit(ncps_statefip), centers = 4, nstart = 25)
str(k2)


fviz_cluster(k2, data = na.omit(ncps_statefip))

na.omit(ncps_statefip) %>%
  mutate(Cluster = k2$cluster) %>%
  group_by(Cluster) %>%
  summarise_all("mean")

na.omit(ncps_statefip) %>%
  mutate(Cluster = k2$cluster) %>%
  group_by(Cluster)

```

The formed clusters are the following:

1) Connecticut, Maryland, New Hampshire
2) Alabama, Arizona, Arkansas, Idaho, Iowa, Maine, Montana, New Mexico, Ohio, Oklahoma, Oregon, South Carolina, Texas, West Virginia
3) California, Georgia, Hawaii, Massachussetts, Minnesota, Missouri, New Jersey, New York, Utah, Virginia, Washington
4) Colorado, Delaware, District of Columbia, Florida, Illinois, Indiana, Kansas, Kentucky, Louisiana, Michigan, Nebraska, North Carolina, North Dakota, Pennsylvania, Tennessee, Wisconsin


# Principal Component Analysis

We are going to perform now a Singular value decomposition exercise.


```{r pca, echo=TRUE,warnings=FALSE,message=FALSE}
ncps_statefip <- aggregate(ncps_sub, by=list(ncps_sub$statefip), FUN=mean, na.rm = TRUE)
ncps_statefip$Group.1 <- NULL
ncps.pca <- prcomp(na.omit(ncps_statefip), center = TRUE,scale. = TRUE)

summary(ncps.pca)
str(ncps.pca)

# Visualize eigenvalues (scree plot). Show the percentage of variances explained by each principal component

fviz_eig(ncps.pca)

# Graph of states States with a similar profile are grouped together

fviz_pca_ind(ncps.pca,
             col.ind = "cos2", # Color by the quality of representation
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE
             )

# Graph of variables. Positive correlated variables point to the same side of the plot. Negative correlated variables point to opposite sides of the graph

fviz_pca_var(ncps.pca,
             col.var = "contrib", # Color by contributions to the PC
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE
             )

groups <- as.factor(ncps_sub$pub[1:dim(ncps_statefip)[1]])

fviz_pca_ind(ncps.pca,
             col.ind = groups, # color by groups
             palette = c("#00AFBB",  "#FC4E07"),
             addEllipses = TRUE, # Concentration ellipses
             ellipse.type = "confidence",
             legend.title = "Groups",
             repel = TRUE
             )

groups <- as.factor(ncps_sub$medicare[1:dim(ncps_statefip)[1]])

fviz_pca_ind(ncps.pca,
             col.ind = groups, # color by groups
             palette = c("#00AFBB",  "#FC4E07"),
             addEllipses = TRUE, # Concentration ellipses
             ellipse.type = "confidence",
             legend.title = "Groups",
             repel = TRUE
             )


```