---
title: "Class 05: Singular Value Decomposition"
output: html_document
---





### The SVD in R

Start by creating some simulated data similar to the ones that were given in
the notes:

```{r}
set.seed(1)
n <- 1e4; p <- 6
X <- matrix(rnorm(n*p), ncol = p)
X[,1] <- X[,1] * (0.01) + X[,2] * (0.99)
b <- c(1,2,3,4,5,6)
epsilon <- rnorm(n)
y <- X %*% b + epsilon
```

And taking the SVD of the matrix `X`:

```{r}
svd_output <- svd(X)
U <- svd_output[["u"]]
V <- svd_output[["v"]]
sigma <- svd_output[["d"]]
```

Verify that the matrix $V^t V$ is equal to the identity (note that it will only
be close the identity because of numerical error):

```{r}
t(V) %*% V
```

Rounding to the 10th decimal place shows the result more clearly:

```{r}
round(t(V) %*% V, 10)
```

Then, run the following code to verify that all of the dimensions are exactly as we
would expect:

```{r}
{
  cat("Dimension of X: ")
  cat(paste(dim(X), collapse = " by "))
  
  cat("\nDimension of U: ") 
  cat(paste(dim(U), collapse = " by "))
  
  cat("\nDimension of V: ")
  cat(paste(dim(V), collapse = " by "))
  
  cat("\nLength of singular values: ")
  cat(length(sigma))
}
```

### Least squares estimator

The following code computes the least squares coefficents using the
matrix $X$ and the response vector $y$:

```{r}
lm.fit(X, y)$coefficients
```

Compute the same equation using the SVD formula for $\beta$:

```{r}
V %*% diag(1 / sigma) %*% t(U) %*% y
```


### Maximum values

I want to simulate the size of $|| X v ||_2^2$ where $v$ is in the unit
ball of dimension 6. Here is the code to do this (just make sure that you
understand what it is doing):

```{r}
N <- 100000                     # number of experiments to run
norms <- rep(0, length=N)       # empty vector to store results in 
for (i in seq_len(N))
{
  v <- runif(6)                 # take a random vector of size 6
  v <- v / sqrt(sum(v^2))       # normalize its length so that it is in the l2 unit ball
  norms[i] <- sum((X %*% v)^2)  # take norm of Xv and store in the object `norms`
}
```

From the theoretical results we would expect the smallest value of norms to be:

```{r}
sigma[length(sigma)]^2
```

And the largest value to be:

```{r}
sigma[1]^2
```

How does this compare with the range of simulated values?

```{r}
range(norms)
```

You will probably find that the upper bound is close to the theoretical
upper bound but the lower bound is a lot larger than the theoretical limit.
Show here that it is in fact possible to achieve the lower bound with the
vector $V[, 6]$:

```{r}
v <- V[, 6]
cat("||v||_2^2  is equal to ", sum(v^2), "\n")
cat("||Xv||_2^2 is equal to ", sum((X %*% v)^2), "\n")
```









