---
title: "Predicting Hospital Exit"
description: |
  Data description 
author:
  - name: Catherine Ishitani
  - name: Gi Kim
    affiliation: Wharton
date: "March 15, 2021"
output:
  distill::distill_article:
    self_contained: true
toc: false 
categories: Catherine and Gi's project
editor_options: 
  chunk_output_type: console
preview: https://patientengagementhit.com/images/sized/rural_hospital-121d13579ca8677ccaacfae5d62f3c8e.png
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

In this post, we wrangle our hospital data a bit and then describe it, focusing on:

1) characteristics of exiting hospitals and

2) elements of our data that may warrant additional consideration in standard ML models. 

## Data
We've added another five years to our data: we now observe the population of U.S. hospitals (~6,300) from 2003-2017. 

Our predictor variables are drawn from five categories: 

- hospital characteristics: such as number of beds, admissions, total and technological services; non-profit & teaching status; Medicaid share; market share; age

- market: population, poverty rate, number of hospitals, and each hospitals' distance from the nearest 100 bed hospital

- ownership structure: whether the hospital is owned by a system or vertically integrated

- financial: revenue, growth, profit margins, uncompensated care costs, cash, debt    

- lagged variables: outcome, ownership, and financial history     

```{r,echo=FALSE}
library(ggplot2)
library(usmap)
library(gganimate)
library(gifski)
library(png)
library(haven)
library(janitor)
library(skimr)
library(dplyr)
library(kableExtra)
library(caret)
library(tidyverse)
library(maptools)
library(rgdal)
library(magick)
library(glue)
library(corrgram)

#our hospital data
hcris <- read_dta("gandhi_hcris_2.dta") %>%  clean_names() %>% remove_empty()
```

## Wrangling

Our data is already tidy, but we spend some time cleaning up the financial variables. HCRIS data are extracted from unaudited financial records submitted to the Centers for Medicare & Medicaid Services (CMS). They're the most complete source of provider financial data but noisy. Prior literature typically winsorizes outliers and unreasonable values. We also create a net debt and financial growth variables.    

```{r,echo=FALSE}
#drop unreasonable values
cleaned <- mutate(hcris,
  liquid     = replace(liquid,(liquid>10)        | (liquid<-1),NA),
  uncomp     = replace(uncomp,(uncomp_mgn>1)     | (uncomp_mgn<-1),NA),
  capex      = replace(capex,(capex_mgn>1)       | (capex_mgn<-1),NA),
  ptnt_mgn   = replace(ptnt_mgn,(ptnt_mgn>1)     | (ptnt_mgn<-1),NA),
  ni_mgn     = replace(ni_mgn,(ni_mgn>1)         | (ni_mgn<-1),NA),
  uncomp_mgn = replace(uncomp_mgn,(uncomp_mgn>1) | (uncomp_mgn<-1),NA),
  levg       = replace(levg,(levg>10)            | (levg<-10),NA),
  capex_mgn  = replace(capex_mgn,(capex_mgn>1)   | (capex_mgn<-1),NA))
#CI: I don't think this code is working.

#create lagged vars (of all outcome & non-fixed predictor variables) 
# if you have the list of variables, you can do this as:
# cleaned <- cleaned %>% mutate(across(c( varlist ), list(lag=~lag(.x))))

```

```{r,echo=TRUE}
#winsorize outliers outside the 1st and 99th percentiles
var_vec = c('ptnt_mgn', 'ni_mgn', 'liquid', 'uncomp', 'uncomp_mgn', 'capex', 'capex_mgn', 'rev_adm', 'tot_assets', 'ptnt_opex', 'oth_costs', 'tot_costs','levg')
clean_p1_p99 <- cleaned %>% group_by(year) %>% summarise_at(vars('ptnt_mgn', 'ni_mgn', 'liquid', 'uncomp', 'uncomp_mgn', 'capex', 'capex_mgn', 'rev_adm', 'tot_assets', 'ptnt_opex', 'oth_costs', 'tot_costs','levg'), .funs=list(p1=~quantile(.,.01,na.rm=TRUE),p99=~quantile(.,.99,na.rm=TRUE)))
clean_p1_p99_merged <- inner_join(cleaned,clean_p1_p99,by='year')

for(varname in var_vec) {
  cleaned$rmflag = (cleaned[{varname}] < clean_p1_p99_merged[glue('{varname}_p1')] | 
                     cleaned[{varname}] > clean_p1_p99_merged[glue('{varname}_p99')] |  
                      is.na(cleaned[{varname}]))
}

cleaned <- cleaned %>% filter(!rmflag)

#create a net debt variable
cleaned <- mutate(hcris, netdebt = debt - cash)

#create growth variables
cleaned <- cleaned %>% group_by(num_prvdr_num) %>% mutate(across(c(rev_tot,rev_netptnt,admtot,net_income,ptnt_income,exptot), list(ch=~(.x-lag(.x))/.x)))

```
