---
title: "Pricing Strategies of Retailers"
description: |
  Final Project Proposal for ECON-712-008
author:
  - name: Zhongtian Chen
  - name: Yihao Yuan
    url: https://github.com/yihao-yuan
    affiliation: The University of Pennsylvania
date: "`r Sys.Date()`"
output: 
  distill::distill_article
toc: false
categories: Zhongtian and Yihao's project
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(lubridate)
library(janitor)
library(dplyr)
library(tidyr)
library(jmuOutlier)
library(Hmisc)
```

## Research Question

Price dispersion has long been central to a range of economic questions. For example, DellaVigna and Gentzkow (2019) investigated how uniform pricing within store chains affect store profits and inequality. However, most papers only focused on price dispersion over stores (e.g., DellaVigna and Gentzkow, 2019) or products (e.g., Hitsch, Hortacsu and Lin, 2021). Few has looked into price dispersion over time, though it is documented that more than 50% of the variations in transaction prices cannot be explained by differences in stores or products (Kaplan and Menzio, 2015). Important exceptions to the lack of studies on price dispersions over time are literatures on intertemporal price discrimination (e.g., Hendel and Nevo, 2013), where stores use cyclical price changes to discriminate between searchers and non-searchers.

In this project, we would like to add to the literatures above by providing patterns of price dispersions over time. Moreover, we want to explain the reasons for a striking pattern presented in our data, that in nearly half of the promotions, the stores discounted prices below their costs. This contradicts the assumptions implied by intertemporal price discrimination, where stores earn positive but quantitatively different profits from different groups of consumers. We plan to investigate other possibilities, such as loss leader strategy and inventory management, to explain the negative margins. In the end, we want to optimize the stores' pricing strategies, and predict profit gains from the new strategies.

## Data

### Data Source

* A large shopping complex in Middle East operated in 15 countries.
* Carrefour data: 66 stores, data from January 2017 to March 2019.
  + Assortment data: product-package-date level; information includes regular price, promotion price, promotion time, quantities sold, quantities in stock, and wholesale price (cost); product-package information only available for some products (e.g., beverages). Data size: 3 billion observations, more than 800 GB.
  + Consumer data: 4.44 million consumers were in their loyalty rewards program; information includes basic demographics (gender and age). 
  + Transaction data: about 29 million transactions made by loyalty program consumers per month across all stores; information includes timestamp, item purchased, price for each item, total spendings. Data size: unknown observations, about 170 GB.
* Supplementary data
  + Credit card data: transactions used by credit card issued by the owner of the shopping complex.
  + Wi-Fi data: search history of consumers when shopping within the complex/Carrefour.
  
### Descriptive Analyses

We start with assortment data on six categories of products: coconut water, corn, cups and straws, deo men roll-on and stick, greek yogurt, and soda and tonic water. We focus on these six categories because the detailed product information of other products are encrypted by the stores. This reduces the number of product-package-days to 19,750,749. We further exclude observations where prices or costs are zero. It further reduces the number of observations to 18,736,746 (94.87%).

```{r}
df <- read.csv('https://www.dropbox.com/s/8kk3nab4s8133z9/assortment_matched.csv?dl=1', 
               header = TRUE, sep = ",")

# A product is on discount if its promotion price is nonzero and is smaller than the regular price
df$discount <- ifelse(df$price_promotion<df$price_regular & df$price_promotion!=0, 1, 0)
df$price <- df$discount*df$price_promotion + (1-df$discount)*df$price_regular
df <- subset(df, cost>0 & price>0)

# Compute Margin = (price-cost)/price
df$margin = (df$price - df$cost)/df$price
summary(df$margin)
df %>%
  group_by(subfamily_name) %>%
  summarise(mean = mean(margin), Q1 = quantile(margin, 0.25), median = median(margin), 
            Q3 = quantile(margin, 0.75), n = n())

df %>%
  group_by(subfamily_name) %>%
  summarise(mean = weighted.mean(margin, sales_quantity), 
            Q1 = wtd.quantile(margin, weights = sales_quantity,probs = 0.25),
            median = wtd.quantile(margin, weights = sales_quantity,probs = 0.5), 
            Q3 = wtd.quantile(margin, weights = sales_quantity,probs = 0.75))
```

The two tables above show the margins of product-packages by categories. The first table presents the summary at product-package-day-level, while second table weights by the number of quantities sold. It turns out that the margins of product-packages with higher sales tend to be higher, as the means and quantiles in the second table are generally higher than in the first table.

```{r, include = FALSE}
sprintf("The share of product-days with negative margins: %f", sum(df$margin < 0)/nrow(df))
sprintf("The share of sales with negative margins: %f", sum((df$margin<0)*df$sales_quantity)/sum(df$sales_quantity))
sprintf("The share of product-days in discount with negative margins: %f", 
        sum(df$margin < 0 & df$discount==1)/sum(df$discount == 1))
sprintf("The share of sales in discount with negative margins: %f", 
        sum((df$margin<0 & df$discount==1)*df$sales_quantity)/sum((df$discount == 1)*df$sales_quantity))
```

A discount is defined as the price lower than the regular price for a product-package. Both daily sales price and regular price are provided in the assortment data. 1.66% of the product-packages-days were in discount, and 7.14% of the product-packages were sold at discounted prices, showing evidence that consumers are price sensitive.

We compute margins of each product-package-day as $Margin = \frac{P-C}{P}$. It turns out that 4.69% (10.56%) of them are sold with negative margins (the number in the parentheses shows the result weighted by sales quantity). Strikingly, 49.48% (63.33%) of the discounted product-packages were sold with negative margins. The histograms of margins for all and discounted product-packages are presented below.

```{r}
truncHist(df$margin, main="Margin rates for all product-packages", xmin=-0.1, xmax=0.4, xlab="Margin rate")
truncHist(df$margin[df$discount==1], main="Margin rates for discounted product-packages", 
          xmin=-1, xmax=0.5,  xlab="Margin rate")
```

## Research Plan

* Read full data. We have not been able to read the full assortment, which is more than 700 GB in size. We are actively working with HPCC to resolve the problem.
* Conduct more detailed descriptive analyses, and nail down the research question.
* Explore the possibilities of some other explanations to negative margins, including but not limited to:
  + Loss leader strategy. Stores may want to offer deep promotions on some products to attract consumers to the store to shop for other products, and maximize the overall profit. We may need to investigate transaction data to confirm whether it is true.
  + Inventory management. Some products are perishable, and stores may want to sell them before they perish. Inventory information is available, but more work is needed to investigate whether inventory management is a reason for deep discounts.

## References 
DellaVigna, S. and Gentzkow, M., 2019. Uniform pricing in us retail chains. *The Quarterly Journal of Economics*, 134(4), pp.2011-2084.

Hendel, I. and Nevo, A., 2013. Intertemporal price discrimination in storable goods markets. *American Economic Review*, 103(7), pp.2722-51.

Hitsch, G.J., Hortacsu, A. and Lin, X., 2021. Prices and promotions in us retail markets: Evidence from big data. *Quantitative Marketing and Economics*, forthcoming.

Kaplan, G. and Menzio, G., 2015. The morphology of price dispersion. *International Economic Review*, 56(4), pp.1165-1206.