---
title: "DoubleML-application"
description: |
  A first attempt in predicting take-up of welfare programs.
author:
  - name: Ornella Darova
  - name: Ozgur Seker
    url: {}
date: 03-30-2021
output:
  distill::distill_article:
    self_contained: false
categories: Ornella and Ozgur's project
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In previous posts, we have prepared our data and set the eligibility variable for different programs. In this post, we will create a setup for double ML package applications. 

```{r prep, echo=FALSE}
library(dplyr)
library(ranger)   
library(rsample)
library(caret)
library(rpart)
library(haven)
library(rlang)

ncps <- read_dta("sampleDataFull.dta")

dataPrep <- function(ncps = ncps){

  
  ### Build eligibility variable for PUBLIC HOUSING
  ncps <- as.data.frame(ncps)
  ncps$hous_elig = ifelse(ncps$famincAssigned<ncps$l80|ncps$disabwrk==2|ncps$age>65,1,0) 
  ncps$hous_elig = ifelse(ncps$disabwrk==0,".",ncps$hous_elig)
  ncps$hous_elig = as.numeric(ncps$hous_elig)
  
  ### Build eligibility variable for MEDICARE
  
  ncps$med_elig = ifelse(ncps$disabwrk==2|ncps$age>65,1,0)
  ncps$med_elig = ifelse(ncps$disabwrk==0,".",ncps$med_elig)
  ncps$med_elig = as.numeric(ncps$med_elig)
  
  
  # Prepare variables
  ncps$pub <- ifelse(ncps$pubhous==1,0,1)
  ncps$pub <- ifelse(ncps$pubhous==0,".",ncps$pub)
  ncps$pub <- as.numeric(ncps$pub)
  
  ncps$medicare <- ifelse(ncps$carely==1,0,1)
  ncps$medicare <- ifelse(ncps$carely==0,".",ncps$medicare)
  ncps$medicare <- as.numeric(ncps$medicare)
  
  ncps$age <- as.numeric(ncps$age)
  ncps$male <- ifelse(ncps$sex==2,0,1)
  
  ncps$male <- ifelse(ncps$sex==9,".",ncps$male)
  ncps$male <- as.numeric(ncps$male)
  
  ncps$black <- ifelse(ncps$race==200,1,".")
  ncps$black <- ifelse(ncps$race==100,0,ncps$black)
  ncps$black <- as.numeric(ncps$black)
  ncps$asian <- ifelse(ncps$race==651,1,0)
  ncps$asian <- as.numeric(ncps$asian)
  
  ncps$married <- ifelse(ncps$marst==1|ncps$marst==2,1,0)
  ncps$married <- ifelse(ncps$marst==9,".",ncps$married)
  ncps$married <- as.numeric(ncps$married)
  
  ncps$foreign <- ifelse(ncps$nativity==4|ncps$nativity==5,1,".")
  ncps$foreign <- ifelse(ncps$nativity==1,0,ncps$foreign)
  ncps$foreign <- as.numeric(ncps$foreign)
  
  ncps$veteran <- ifelse(ncps$vetstat==1,0,".")
  ncps$veteran <- ifelse(ncps$vetstat==2,1,ncps$veteran)
  ncps$veteran <- as.numeric(ncps$veteran)
  
  ncps$nchildren <- as.numeric(ncps$nchild)
  
  ncps$unemployed <- ifelse(ncps$labforce==2,0,1)
  ncps$unemployed <- ifelse(ncps$labforce==0,".",ncps$unemployed)
  ncps$unemployed <- as.numeric(ncps$unemployed)
  
  return(ncps)
}

ncps <- dataPrep(ncps)


```

```{r packages, echo=TRUE}
library(DoubleML)
library(mlr3learners)
library(mlr3)
library(data.table)
library(randomForest)
```

The function below prepares data and apply doubleML with random forest. If argument subset is true, it uses the subset of data according to given arguments. 

```{r packages2, echo=TRUE}
effectByDML <- function(data = ncps, yvar, dvar, xvars, subset=FALSE, subsetcols = NA, subsetvars = NA){
  df <- zap_labels( data )
  if(subset){
    K <- length(subsetcols)
    for (i in 1:K){
      df <- df %>% filter(!!as.name(subsetcols[i]) == subsetvars[i])
    }
  }
  
  x_cols <- xvars[xvars != dvar]

  
  df <- subset(df, select = c(yvar, dvar, xvars))
  df <- df[complete.cases(df),]
  df <- df %>% mutate_at(yvar, as.numeric)
  

  # Estimating the PLR
  lgr::get_logger("mlr3")$set_threshold("warn")
  
  learner = lrn("regr.ranger", num.trees = 100, mtry = 3, min.node.size = 2, max.depth = 5,
                respect.unordered.factors = 'partition')
  ml_g = learner$clone()
  ml_m = learner$clone()
  set.seed(1111)
  obj_dml_data = double_ml_data_from_data_frame(df,
                                                y_col = yvar,
                                                d_cols = dvar,
                                                x_cols = x_cols,
                                                         use_other_treat_as_covariate=TRUE)
  dml_plr_obj = DoubleMLPLR$new(obj_dml_data, ml_g, ml_m)
  dml_plr_obj$fit()
  return(dml_plr_obj)
}


```

Before running the code, we detect ordered variables in our dataset and make all other variables categorical variable. 

```{r prepare, echo=TRUE}
orderedVars <- c("heatval", "stampmo", "atelunch", "frelunch", "faminc", "unitsstr", 
                 "age", "famsize", "nchild", "nchlt5", "famunit", "eldch", "yngch", "durunem2",
                 "uhrsworkt", "educ", "firmsize", "health", "famincAssigned", "hhincome", "income_county")

ncps <- ncps %>% mutate_all( as.factor)
ncps <- ncps %>% mutate_at(orderedVars, as.numeric)


```

We set our output interest as y, variables we want to estimate their effects as dlist, all covariates as xvars, and the variables that their effects should be normalized as dlistNormalize. 

```{r setting, echo=TRUE}
y <- c("pub", "rentsub", "heatsub", "foodstmp", "lunchsub", "caidly", "carely")
dlist <- c("famsize", "famincAssigned", "income_county", "nchild", "phone", "black", "vetstat", "citizen", "nativity", "empstat", "union")
dlistNormalize <- c("famincAssigned", "income_county")
xvars <- c("famsize","famincAssigned", "age", "empstat", "labforce", "famunit", "firmsize", "health", "income_county","phinsur", "nchild", "black", "disabwrk", "sex", "vetstat", "citizen", "nativity", "occupation", "empstat")
```

We are interested in estimating effect on take-up rate for all and for only eligible people. Difference between these two might also give us some intuition. The code below prepares a table for results. 


```{r running, echo=TRUE}

resultsObjectsAll <- list()
resultsObjectsElig <- list()
resultsDF <- c()

for (i in 1:length(dlist)) {
  d <- dlist[i]
  print(paste0("Calculating effect of ", d, " on take up rate for all"))
  resAllObj  <- effectByDML(data = ncps, yvar = y, dvar = d, xvars = xvars, subset = FALSE, subsetcols = "hous_elig", subsetvars = 1)
  if (d %in% dlistNormalize){
    resAll  <- c(resAllObj$coef*sqrt(var(ncps[,d], na.rm = TRUE)), resAllObj$t_stat)
    resAll  <- round(resAll, 6)
  } else {
    resAll  <- c(resAllObj$coef, resAllObj$t_stat)
    resAll  <- round(resAll, 6)
  }
  resultsObjectsAll[[i]] <- resAllObj
  
  print(paste0("Calculating effect of ", d, " on take up rate for eligible only"))
  resEligObj <- effectByDML(data = ncps, yvar = y, dvar = d, xvars = xvars, subset = TRUE, subsetcols = "hous_elig", subsetvars = 1)
  
  if (d %in% dlistNormalize){
    resElig  <- c(resEligObj$coef*sqrt(var(ncps[,d], na.rm = TRUE)), resEligObj$t_stat)
    resElig  <- round(resElig, 6)
  } else {
    resElig  <- c(resEligObj$coef, resEligObj$t_stat)
    resElig  <- round(resElig, 6)
  }
  
  resultsObjectsElig[[i]] <- resEligObj
  
  resultsDF <- rbind(resultsDF, c(d, resAll, resElig))
}

colnames(resultsDF) <- c("d_var", "coef_all", "t_stat_all", "coef_elig", "t_stat_elig")
print(resultsDF)
```
