---
title: "DoubleML-application"
description: |
  A first attempt in predicting take-up of welfare programs.
author:
  - name: Ornella Darova
  - name: Ozgur Seker
    url: {}
date: "`r Sys.Date()`"
output: distill::distill_article
categories: Ornella and Ozgur's project
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# root = "../../../Predicting Take-Up Rate/"
```

# Introduction

In previous posts, we prepared our data and set the eligibility variables for different programs. We also started inspecting possible predictors of the take-up of welfare programs through random forests. 

In this post, we will create a setup for double ML package applications. 

```{r prep, echo=FALSE}
library(dplyr)
library(ranger)   
library(rsample)
library(caret)
library(rpart)
library(haven)
library(rlang)
library(tibble)

ncps <- read_dta("sampleDataFull.dta")

dataPrep <- function(ncps = ncps){

  
  ### Build eligibility variable for PUBLIC HOUSING
  ncps <- as.data.frame(ncps)
  ncps$hous_elig = ifelse(ncps$famincAssigned<ncps$l80|ncps$disabwrk==2|ncps$age>65,1,0) 
  ncps$hous_elig = ifelse(ncps$disabwrk==0,".",ncps$hous_elig)
  ncps$hous_elig = as.numeric(ncps$hous_elig)
  
  ### Build eligibility variable for MEDICARE
  
  ncps$med_elig = ifelse(ncps$disabwrk==2|ncps$age>65,1,0)
  ncps$med_elig = ifelse(ncps$disabwrk==0,".",ncps$med_elig)
  ncps$med_elig = as.numeric(ncps$med_elig)
  
  
  # Prepare variables
  ncps$pub <- ifelse(ncps$pubhous==1,0,1)
  ncps$pub <- ifelse(ncps$pubhous==0,".",ncps$pub)
  ncps$pub <- as.numeric(ncps$pub)
  
  ncps$medicare <- ifelse(ncps$carely==1,0,1)
  ncps$medicare <- ifelse(ncps$carely==0,".",ncps$medicare)
  ncps$medicare <- as.numeric(ncps$medicare)
  
  ncps$age <- as.numeric(ncps$age)
  ncps$male <- ifelse(ncps$sex==2,0,1)
  
  ncps$male <- ifelse(ncps$sex==9,".",ncps$male)
  ncps$male <- as.numeric(ncps$male)
  
  ncps$black <- ifelse(ncps$race==200,1,".")
  ncps$black <- ifelse(ncps$race==100,0,ncps$black)
  ncps$black <- as.numeric(ncps$black)
  ncps$asian <- ifelse(ncps$race==651,1,0)
  ncps$asian <- as.numeric(ncps$asian)
  
  ncps$married <- ifelse(ncps$marst==1|ncps$marst==2,1,0)
  ncps$married <- ifelse(ncps$marst==9,".",ncps$married)
  ncps$married <- as.numeric(ncps$married)
  
  ncps$foreign <- ifelse(ncps$nativity==4|ncps$nativity==5,1,".")
  ncps$foreign <- ifelse(ncps$nativity==1,0,ncps$foreign)
  ncps$foreign <- as.numeric(ncps$foreign)
  
  ncps$veteran <- ifelse(ncps$vetstat==1,0,".")
  ncps$veteran <- ifelse(ncps$vetstat==2,1,ncps$veteran)
  ncps$veteran <- as.numeric(ncps$veteran)
  
  ncps$nchildren <- as.numeric(ncps$nchild)
  
  ncps$unemployed <- ifelse(ncps$labforce==2,0,1)
  ncps$unemployed <- ifelse(ncps$labforce==0,".",ncps$unemployed)
  ncps$unemployed <- as.numeric(ncps$unemployed)
  
  return(ncps)
}

ncps <- dataPrep(ncps)


```

```{r packages, echo=TRUE}
library(DoubleML)
library(mlr3learners)
library(mlr3)
library(data.table)
library(randomForest)
```

# DoubleML

The function below prepares data and applies doubleML with random forest. If the argument subset is true, it uses the subset of data according to given arguments. 

```{r function, echo=TRUE}
effectByDML <- function(data = ncps, yvar, dvar, xvars, subset=FALSE, subsetcols = NA, subsetvars = NA){
  df <- zap_labels( data )
  if(subset){
    K <- length(subsetcols)
    for (i in 1:K){
      df <- df %>% filter(!!as.name(subsetcols[i]) == subsetvars[i])
    }
  }
  
  x_cols <- xvars[xvars != dvar]

  
  df <- subset(df, select = c(yvar, dvar, xvars))
  df <- df[complete.cases(df),]
  df <- df %>% mutate_at(yvar, as.numeric)
  

  # Estimating the PLR
  lgr::get_logger("mlr3")$set_threshold("warn")
  
  learner = lrn("regr.ranger", num.trees = 100, mtry = 3, min.node.size = 2, max.depth = 5,
                respect.unordered.factors = 'partition')
  ml_g = learner$clone()
  ml_m = learner$clone()
  set.seed(1111)
  obj_dml_data = double_ml_data_from_data_frame(df,
                                                y_col = yvar,
                                                d_cols = dvar,
                                                x_cols = x_cols,
                                                         use_other_treat_as_covariate=TRUE)
  dml_plr_obj = DoubleMLPLR$new(obj_dml_data, ml_g, ml_m)
  dml_plr_obj$fit()
  return(dml_plr_obj)
}


```

Before running the code, we detect ordered variables in our dataset and convert all categorical variables to factor variables. 

```{r prepare, echo=TRUE}
orderedVars <- c("heatval", "stampmo", "atelunch", "frelunch", "faminc", "unitsstr", 
                 "age", "famsize", "nchild", "nchlt5", "famunit", "eldch", "yngch", "durunem2",
                 "uhrsworkt", "educ", "firmsize", "health", "famincAssigned", "hhincome", 
                 "income_county", "phone", "black", "vetstat", "citizen", "nativity", "empstat", "union")

ncps <- ncps %>% mutate_all(as.factor)
ncps <- ncps %>% mutate_at(orderedVars, as.numeric)


```

We set our output interest as y, variables we want to estimate their effects as dlist, all covariates as xvars, and the variables that their effects should be normalized as dlistNormalize. 

## Public housing

```{r setting, echo=TRUE}
y <- "pub"
dlist <- c("famsize", "famincAssigned", "income_county", "nchild", "phone", "black", "vetstat", "citizen", "nativity", "empstat", "union")
dlistNormalize <- c("famincAssigned", "income_county")
xvars <- c("famsize","famincAssigned", "age", "empstat", "labforce", "famunit", "firmsize", "health", "income_county", "phinsur", "nchild", "black", "disabwrk", "sex")
```

We are interested in estimating effect on take-up rate of public housing for all, and subsequently exclusively for eligible people. The code below prepares a table for results. 


```{r running, echo=TRUE}

resultsObjectsAll <- list()
resultsObjectsElig <- list()
resultsDF <- c()

for (i in 1:length(dlist)) {
  d <- dlist[i]
  print(paste0("Calculating effect of ", d, " on take up rate for all"))
  resAllObj  <- effectByDML(data = ncps, yvar = y, dvar = d, xvars = xvars, subset = FALSE, subsetcols = "hous_elig", subsetvars = 1)
  if (d %in% dlistNormalize){
    resAll  <- c(resAllObj$coef*sqrt(var(ncps[,d], na.rm = TRUE)), resAllObj$t_stat)
    resAll  <- round(resAll, 6)
  } else {
    resAll  <- c(resAllObj$coef, resAllObj$t_stat)
    resAll  <- round(resAll, 6)
  }
  resultsObjectsAll[[i]] <- resAllObj
  
  print(paste0("Calculating effect of ", d, " on take up rate for eligible only"))
  resEligObj <- effectByDML(data = ncps, yvar = y, dvar = d, xvars = xvars, subset = TRUE, subsetcols = "hous_elig", subsetvars = 1)
  
  if (d %in% dlistNormalize){
    resElig  <- c(resEligObj$coef*sqrt(var(ncps[,d], na.rm = TRUE)), resEligObj$t_stat)
    resElig  <- round(resElig, 6)
  } else {
    resElig  <- c(resEligObj$coef, resEligObj$t_stat)
    resElig  <- round(resElig, 6)
  }
  
  resultsObjectsElig[[i]] <- resEligObj
  
  resultsDF <- rbind(resultsDF, c(d, resAll, resElig))
}

colnames(resultsDF) <- c("d_var", "coef_all", "t_stat_all", "coef_elig", "t_stat_elig")
print.data.frame(as.data.frame( resultsDF) )
```

We see that income per capita of the county has a positive effect on the take-up rates in full data: this is due to the fact that when per-capita income is high, more people are eligible due to increasing poverty levels, as eligibility is set in terms of a percentage of the median income in the corresponding county. When we run the same empirical strategy exclusively for the pool of eligible individuals, we observe that this effect, indeed, becomes  not significant.

The family income still represents an important factor for the take-up rate for the pool of eligible individuals. 

We can observe that, when we remove the number of children from the covariates, we suddenly start to see a significant effect of family size on take-up rates. However, it is not significant when the number of children is included. On the other hand, even though family size is included in covariates, effect of the number of children is significant. 

Correlation between family size and number of children is about 0.6, so we could expect the 60% of the effect of nchild as family size effect when we remove nchild. However the effect estimated with doubleML is much lower than that. 

```{r running2, echo=FALSE}
# y <- "pub"
# dlist <- c("famsize")
# dlistNormalize <- c("famincAssigned", "income_county")
# xvars <- c("famsize","famincAssigned", "age", "empstat", "labforce", "famunit", "firmsize", "health", "income_county","phinsur", "black", "disabwrk", "sex")

y <- "pub"
dlist <- c("famsize")
dlistNormalize <- c("famincAssigned", "income_county")
xvars <- c("famsize","famincAssigned", "age", "empstat", "labforce", "famunit", "firmsize", "health", "income_county", "phinsur", "black", "disabwrk", "sex")

resultsObjectsAlln <- list()
resultsObjectsElign <- list()
resultsDF <- c()

for (i in 1:length(dlist)) {
  d <- dlist[i]
  # print(paste0("Calculating effect of ", d, " on take up rate for all"))
  resAllObj  <- effectByDML(data = ncps, yvar = y, dvar = d, xvars = xvars, subset = FALSE, subsetcols = "hous_elig", subsetvars = 1)
  if (d %in% dlistNormalize){
    resAll  <- c(resAllObj$coef*sqrt(var(ncps[,d], na.rm = TRUE)), resAllObj$t_stat)
    resAll  <- round(resAll, 6)
  } else {
    resAll  <- c(resAllObj$coef, resAllObj$t_stat)
    resAll  <- round(resAll, 6)
  }
  resultsObjectsAlln[[i]] <- resAllObj
  
  # print(paste0("Calculating effect of ", d, " on take up rate for eligible only"))
  resEligObj <- effectByDML(data = ncps, yvar = y, dvar = d, xvars = xvars, subset = TRUE, subsetcols = "hous_elig", subsetvars = 1)
  
  if (d %in% dlistNormalize){
    resElig  <- c(resEligObj$coef*sqrt(var(ncps[,d], na.rm = TRUE)), resEligObj$t_stat)
    resElig  <- round(resElig, 6)
  } else {
    resElig  <- c(resEligObj$coef, resEligObj$t_stat)
    resElig  <- round(resElig, 6)
  }
  
  resultsObjectsElign[[i]] <- resEligObj
  
  resultsDF <- rbind(resultsDF, c(d, resAll, resElig))
}

colnames(resultsDF) <- c("d_var", "coef_all", "t_stat_all", "coef_elig", "t_stat_elig")
print.data.frame(as.data.frame( resultsDF) )
```

We also wanted to check effects of the number of children, family size, and the variable nchlt5 (number of children that are younger than 5) when nchlt is also included. 

```{r running3, echo=FALSE}
y <- "pub"
dlist <- c("famsize","nchild", "nchlt5")
dlistNormalize <- c("famincAssigned", "income_county")
xvars <- c("famsize","famincAssigned", "age", "empstat", "labforce", "famunit", "firmsize", "health", "income_county","phinsur", "black", "disabwrk", "sex", "nchild", "nchlt5")

resultsObjectsAlln <- list()
resultsObjectsElign <- list()
resultsDF <- c()

for (i in 1:length(dlist)) {
  d <- dlist[i]
  # print(paste0("Calculating effect of ", d, " on take up rate for all"))
  resAllObj  <- effectByDML(data = ncps, yvar = y, dvar = d, xvars = xvars, subset = FALSE, subsetcols = "hous_elig", subsetvars = 1)
  if (d %in% dlistNormalize){
    resAll  <- c(resAllObj$coef*sqrt(var(ncps[,d], na.rm = TRUE)), resAllObj$t_stat)
    resAll  <- round(resAll, 6)
  } else {
    resAll  <- c(resAllObj$coef, resAllObj$t_stat)
    resAll  <- round(resAll, 6)
  }
  resultsObjectsAlln[[i]] <- resAllObj
  
  # print(paste0("Calculating effect of ", d, " on take up rate for eligible only"))
  resEligObj <- effectByDML(data = ncps, yvar = y, dvar = d, xvars = xvars, subset = TRUE, subsetcols = "hous_elig", subsetvars = 1)
  
  if (d %in% dlistNormalize){
    resElig  <- c(resEligObj$coef*sqrt(var(ncps[,d], na.rm = TRUE)), resEligObj$t_stat)
    resElig  <- round(resElig, 6)
  } else {
    resElig  <- c(resEligObj$coef, resEligObj$t_stat)
    resElig  <- round(resElig, 6)
  }
  
  resultsObjectsElign[[i]] <- resEligObj
  
  resultsDF <- rbind(resultsDF, c(d, resAll, resElig))
}

colnames(resultsDF) <- c("d_var", "coef_all", "t_stat_all", "coef_elig", "t_stat_elig")
print.data.frame(as.data.frame( resultsDF) )
```

After including nchlt5, the significance of the family size decreases more, while we can observe that the variable nchlt5 has a significant effect on take-up rates, but it is lower than the number of all children. In other words: the number of children seems to have a strong effect, and the younger children drive a large part of this impact, while once taken into consideration the number of children, the effect of family size per se is not that relevant. This impaact could come from several factors: first of all, it could be the case that households with a higher number of very young children are given priority in these programs; moreover, it could be the case that households with very young children are more easily approached by several institutions that might inform them: hospitals, kindergartens and so on. 










