---
title: "Learning Approach"
description: |
  An attempt to approach social programs as a learning problem
author:
  - name: Ozgur Seker
    url: {}
  - name: Ornella Darova
    url: {}
date: 05-03-2021
output:
  distill::distill_article:
    self_contained: false
categories: Ornella and Ozgur's project
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(cache = TRUE)
```

```{r prep, echo=FALSE, warning=FALSE}
library(dplyr)
library(ranger)   
library(rsample)
library(caret)
library(rpart)
library(haven)
library(rlang)
library(tibble)

ncps <- read_dta("sampleDataFull.dta")

dataPrep <- function(ncps = ncps){

  
  ### Build eligibility variable for PUBLIC HOUSING
  ncps <- as.data.frame(ncps)
  ncps$hous_elig = ifelse(ncps$famincAssigned<ncps$l80|ncps$disabwrk==2|ncps$age>65,1,0) 
  ncps$hous_elig = ifelse(ncps$disabwrk==0,".",ncps$hous_elig)
  ncps$hous_elig = as.numeric(ncps$hous_elig)
  
  ### Build eligibility variable for MEDICARE
  
  ncps$med_elig = ifelse(ncps$disabwrk==2|ncps$age>65,1,0)
  ncps$med_elig = ifelse(ncps$disabwrk==0,".",ncps$med_elig)
  ncps$med_elig = as.numeric(ncps$med_elig)
  
  
  # Prepare variables
  ncps$pub <- ifelse(ncps$pubhous==1,0,1)
  ncps$pub <- ifelse(ncps$pubhous==0,".",ncps$pub)
  ncps$pub <- as.numeric(ncps$pub)
  
  ncps$medicare <- ifelse(ncps$carely==1,0,1)
  ncps$medicare <- ifelse(ncps$carely==0,".",ncps$medicare)
  ncps$medicare <- as.numeric(ncps$medicare)
  
  ncps$age <- as.numeric(ncps$age)
  ncps$male <- ifelse(ncps$sex==2,0,1)
  
  ncps$male <- ifelse(ncps$sex==9,".",ncps$male)
  ncps$male <- as.numeric(ncps$male)
  
  ncps$black <- ifelse(ncps$race==200,1,".")
  ncps$black <- ifelse(ncps$race==100,0,ncps$black)
  ncps$black <- as.numeric(ncps$black)
  ncps$asian <- ifelse(ncps$race==651,1,0)
  ncps$asian <- as.numeric(ncps$asian)
  
  ncps$married <- ifelse(ncps$marst==1|ncps$marst==2,1,0)
  ncps$married <- ifelse(ncps$marst==9,".",ncps$married)
  ncps$married <- as.numeric(ncps$married)
  
  ncps$foreign <- ifelse(ncps$nativity==4|ncps$nativity==5,1,".")
  ncps$foreign <- ifelse(ncps$nativity==1,0,ncps$foreign)
  ncps$foreign <- as.numeric(ncps$foreign)
  
  ncps$veteran <- ifelse(ncps$vetstat==1,0,".")
  ncps$veteran <- ifelse(ncps$vetstat==2,1,ncps$veteran)
  ncps$veteran <- as.numeric(ncps$veteran)
  
  ncps$nchildren <- as.numeric(ncps$nchild)
  
  ncps$unemployed <- ifelse(ncps$labforce==2,0,1)
  ncps$unemployed <- ifelse(ncps$labforce==0,".",ncps$unemployed)
  ncps$unemployed <- as.numeric(ncps$unemployed)
  
  return(ncps)
}

ncps <- dataPrep(ncps)


```

One of the important observations we have from the data is we have people having social programs even they are not eligible according to the online program descriptions.

```{r notelig, echo= TRUE}

round(mean(ncps[ncps$hous_elig==0,"pub"] , na.rm = TRUE ), 4)
round(mean(ncps[ncps$med_elig==0,"medicare"] , na.rm = TRUE ), 4)
```

Here we could see that almost 1 percent of not eligible people benefit from public houses and 2 percent of not eligible people benefit from medicare. If we think that small portion of not eligible people are applying for these programs, these ratios become much more important. 


So we could conclude that there is no strict boundaries on who will be approved for the social programs and it shouldn't be. 

We propose a model to understand approval process and trade-offs inspired by learning processes. An analogy to understand trade-offs is below. 

Assume that you have arrived a school with treats and you are waiting for children in the garden to give them treats. You like some children more than others but you can choose only give a treat or not when a child comes to you. You can't choose who will want a treat from you. The main trade-off is that, you want to keep treats for your favorable children. However, if you give to an unfavorable children, he/she will inform other children about you and more favorable children will come. 

Similarly, officials are facing a budget constraint while accepting for social programs. On the other hand, depending on the characteristics of the applicant, they have reward from accepting the applicant. Up to now, it is standart budget constraint. In addition to this, each approved applicant will encourage people in the same and/or lower income tier to apply and each rejected will discourage people in the same and/or higher income tier to apply. 

This second effect creates the dynamic state transitions we need for learning. 

So we could formulate our process as following:

**State:**

- Remaining budget/capacity
- Number of applicants and their characteristics
  
**Action:**

- Reject/Approve applicants
  
**Reward:**

- Rewards from approved applicants are function of their characteristics decided by policy makers. 

  
**State Transition: **

- Number of future applicants and their distribution in terms of characteristics are probabilistically dependent on past approval/rejection decisions. 
- This dependence is a black box now but a structural model could be suggested with modeling interactions.
- This part could be estimated if we have the track of applications with their date and approval status. 

  




